---
title: "Satellite Launch Failure"
output: word_document
date: "2024-10-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Probability of a Satellite Launch Failure

First we load the data fream and create a faults table.

```{r, results='hide'}
# Computer faults dataframe
library(dplyr)
faults <- read.csv("C:\\Users\\jeff\\OneDrive\\Books\\Prob and Stats\\Data\\comp_faults.csv")
faults %>%
  count(computer_number, fault, sort=T)

# Creating two-way table from data frame
Faults_Table <- addmargins(table("Fault"=faults$computer_number,
                                         "Frequency"=faults$fault))
```

Here we view the faults frequency table.

```{r}
# View table
Faults_Table
```

Next, we create the probability space and calculate the conditional probabilties and the probabilities of independent events.

```{r, results='hide'}
library(probs)

# Create Probabilty Space
Faults_Data <- probspace(faults)
Faults_Data

# Conditional Probabilities
Cumputer1_fault_prob <- Prob(Faults_Data, event=computer_number == "1", given=fault == "1")
Cumputer1_fault_prob
Cumputer1_nofault_prob <- 1-Cumputer1_fault_prob
Cumputer1_nofault_prob

Cumputer2_fault_prob <- Prob(Faults_Data, event=computer_number == "2", given=fault == "1")
Cumputer2_fault_prob
Cumputer2_nofault_prob <- 1-Cumputer1_fault_prob
Cumputer1_nofault_prob

Cumputer3_fault_prob <- Prob(Faults_Data, event=computer_number == "3", given=fault == "1")
Cumputer3_fault_prob
Cumputer3_nofault_prob <- 1-Cumputer1_fault_prob
Cumputer3_nofault_prob

# Independent Event Probabilties
Prob_111 <-Cumputer1_fault_prob*Cumputer2_fault_prob*Cumputer3_fault_prob
Prob_110 <-Cumputer1_fault_prob*Cumputer2_fault_prob*Cumputer3_nofault_prob
Prob_101 <-Cumputer1_fault_prob*Cumputer2_nofault_prob*Cumputer3_fault_prob
Prob_100 <-Cumputer1_fault_prob*Cumputer2_nofault_prob*Cumputer3_nofault_prob

Prob_011 <-Cumputer1_nofault_prob*Cumputer2_fault_prob*Cumputer3_fault_prob
Prob_010 <-Cumputer1_nofault_prob*Cumputer2_fault_prob*Cumputer3_nofault_prob
Prob_001 <-Cumputer1_nofault_prob*Cumputer2_nofault_prob*Cumputer3_fault_prob
Prob_000 <-Cumputer1_nofault_prob*Cumputer2_nofault_prob*Cumputer3_nofault_prob

# Probaility of computers in use
print(paste0("Computer1 fails, Computer2 fails, Computer3 fails =", round(Prob_111,8)))
print(paste0("Computer1 fails, Computer2 fails, Computer3 works =", round(Prob_110,8)))
print(paste0("Computer1 fails, Computer2 works, Computer3 fails =", round(Prob_101,8)))      
print(paste0("Computer1 fails, Computer2 works, Computer3 works =", round(Prob_100,8)))
print(paste0("Computer1 works, Computer2 fails, Computer3 fails =", round(Prob_011,8)))
print(paste0("Computer1 works, Computer2 fails, Computer3 works =", round(Prob_010,8)))
print(paste0("Computer1 works, Computer2 works, Computer3 fails =", round(Prob_001,8)))
print(paste0("Computer1 works, Computer2 works, Computer3 works =", round(Prob_000,8)))


```

Now, we compute the probability that a satellie launch will fails, i.e., all threee computers fail.

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the result.

```{r echo=FALSE}
# Probability of a system failure
print(paste0("################## RESULT ######################"))
print(paste0("#  Probability of a system failure =", round(Prob_111,8),"  #"))
print(paste0("################################################"))
```